### 2021-07-05 rad-icfp-2021/background.tex
minor syntax

The reader may also notice that $h$ encodes all the partial derivatives of
$g$. For example, 
if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
$h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
and similarly 
$h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
And conversely, if $g$ is differentiable in each argument, then
a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
combinations of partial derivatives:
\[\textstyle h(x_1,x_1',x_2,x_2')=(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))\text.\]

@@

The problem with the previous approach shows up when one wants to compute the full gradient of a function $\RR^n\to\RR$, for a large $n$. 
Forward-mode only computes one directional derivative, for instance one partial derivative. This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
Suppose given a function $f=f_n\circ...\circ f_1 :\RR^n\to\RR$. 
The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(...(Jf_1v))...)$ for a direction $v\in\RR^n$. 
Reverse-mode, on the other hand, computes $(Jf)^Tv =J^Tf_1(J^Tf_{2}(...(J^Tf_nv))...)$ for a vector $v\in\RR$.
In particular, taking $v=1$ computes the gradient of $f$.

@@

Let the term we want to differentiate be $x1:\RR,...,xn:\RR\vdash exp(cos(xi))$.
Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
However, by simply doing this we lose compositionality. 
So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
Similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ...\circ f_1$. 
If the original program was of type $\RR$, then the return value of the original program is the last component of $f$.
Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
\[\nabla f= J^Tf(0,...,0,1)=J^Tf_1(J^Tf_{2}(...(J^Tf_m(0,...,0,1)))...)\]

To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation.
$f_i$ is turned into $\Dsynrevsymbol{f_i}:=<f_i, \lambda Y. Y\circ J^Tf_i>$ where $Y:\RR^{n+i-1}\to \RR^n$. 
We recover compositionality by noting that $<f_{i+1}(f_i), (\lambda Y. Y\circ J^Tf_{i+1}>)(\lambda Y. Y\circ J^Tf_i)>$ reduces to
$<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,...,0,1)$, 
we have obtained a purely functional to compute $\nabla f$.

@@

If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
the row is zero except for at most two indices (one for unary operators).
This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,...,y(n+i)).(y1,...,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
where $[k]$ means that the element is added at the $k$-th index of the tuple.

>>>>

The reader may also notice that $h$ encodes all the partial derivatives of
$g$. For example, 
if $g \colon \RR^2\to \RR$, then with $f_1(x)\defeq x$ and $f_2(x)\defeq x_2$, by applying \eqref{eqn:dualnumber} to $x_1$ we obtain
$h(x_1,1,x_2,0)=(g(x_1,x_2), \frac {\partial g(x,x_2)}{\partial x}(x_1))$
and similarly 
$h(x_1,0,x_2,1)=(g(x_1,x_2), \frac {\partial g(x_1,x)}{\partial x}(x_2))$.
And conversely, if $g$ is differentiable in each argument, then
a unique $h$ satisfying \eqref{eqn:dualnumber} can be found by taking linear
combinations of partial derivatives:

\begin{center}
    \begin{tabular}{r c l}
        $h(x_1,x_1',x_2,x_2')$ &= & $(g(x_1,x_2),x_1' \cdot\frac {\partial g(x,x_2)}{\partial x}(x_1)+x_2'\cdot \frac {\partial g(x_1,x)}{\partial x}(x_2))$
    \end{tabular}
\end{center}

@@

The problem with the previous approach shows up when one wants to compute the full gradient of a function $\RR^n\to\RR$, for a large $n$. 
Forward-mode only computes one directional derivative, for instance one partial derivative. This implies $n$ passes must be performed through the forward derivative to compute the whole gradient.
By using the symmetry in the chain rule, there is a way to compute the whole gradient faster, and this method is reverse-mode differentiation.
Suppose given a function $f=f_n\circ\ldots\circ f_1 :\RR^n\to\RR$. 
The simplest way to see this mathematically, is first to say that forward mode essentially computes $(Jf)v=Jf_n(Jf_{n-1}(\ldots(Jf_1v))\ldots)$ for a direction $v\in\RR^n$. 
Reverse-mode, on the other hand, computes $(Jf)^Tv =J^Tf_1(J^Tf_{2}(\ldots(J^Tf_nv))\ldots)$ for a vector $v\in\RR$.
In particular, taking $v=1$ computes the gradient of $f$.

@@

Let the term we want to differentiate be $x1:\RR,\ldots,xn:\RR\vdash exp(cos(xi))$.
Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
However, by simply doing this we lose compositionality. 
So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
Similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ\ldots\circ f_1$. 
If the original program was of type $\RR$, then the return value of the original program is the last component of $f$.
Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
\begin{center}
    \begin{tabular}{r c l}
        $\nabla f$ &=& $J^Tf(0,\ldots,0,1)=J^Tf_1(J^Tf_{2}(\ldots(J^Tf_m(0,\ldots,0,1))\ldots)$
    \end{tabular}
\end{center}

To actually reverse the order of computation needed for this transpose of jacobians, we use a sort of simple continuation.
$f_i$ is turned into $\Dsynrevsymbol{f_i}:=<f_i, \lambda Y. Y\circ J^Tf_i>$ where $Y:\RR^{n+i-1}\to \RR^n$. 
We recover compositionality by noting that $<f_{i+1}(f_i), (\lambda Y. Y\circ J^Tf_{i+1}>)(\lambda Y. Y\circ J^Tf_i)>$ reduces to
$<f_{i+1}\circ f_i, \lambda Y. Y\circ J^Tf_i \circ J^Tf_{i+1}>$, and thus by induction we can obtain $<f, \lambda Y. YJ^Tf>$.
By applying the identity continuation $\RR^n\to\RR^n$ on the second component then the result to $(0,\ldots,0,1)$, 
we have obtained a purely functional to compute $\nabla f$.

@@

If we look at $J^Tf_{i+1}$, we notice that this function is almost the identity, except at the last row. 
Even on the last row, if the original term was a unary or binary operator like $cos, exp, +, *$, 
the row is zero except for at most two indices (one for unary operators).
This means we can use a more compact representation $J^T\sem{op(xi,xj)}:=\lambda (y1,\ldots,y(n+i)).(y1,\ldots,y(n+i))+[i]\partial_1op(xi,xj)+[j]\partial_2op(xi,xj)$, 
where $[k]$ means that the element is added at the $k$-th index of the tuple.

### 2021-07-05 rad-icfp-2021/correctness.tex
minor syntax

\begin{definition}[Mutlicat from concat]
    A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
    with morphisms $A\to [B1,...,Bn]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
\end{definition}

>>>>

\begin{definition}[Mutlicat from concat]
    A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
    with morphisms $A\to [B_1,\ldots,B_n]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
\end{definition}

### 2021-07-05 rad-icfp-2021/generalization.tex
fixing array notation
minor
minor syntax

We can now compute the reverse-mode transformation as follows.
\begin{center}
\begin{tabular}{r c l}
$\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
&=& ... as before up until !A3! \\
&& !let B0 = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A0 A) in!\\
&& !let B1 = scanr !$1_\Gamma$ $\widehat{\times}$ !B0 in! \\
&& !let B2 = reduce !$\widehat{+}$ $0_\Gamma$! B1 in! \\
&& !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
&& !Y((x1,...,xn)!$\widehat{+}$!B2,0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

@@

Finally, the former point implies we need a few more array manipulations. 
We write !c::[v1,...,vn]! to be the array ![c,v1,...,vn]!. 
Given an array ![v1,...,vn]! of size $n>0$, we write !let x,A=[v1,...,vn]! to mean 
that !x=v1! and !A=[v2,...,vn]!. The reverse derivative of fold left is then
\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e1) e2 e3!) &=&
            !let v,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
            && !let A,Y2 = !$\directD{\rho}{\Gamma,v}{Y1}$!(e3) in! \\
            && !let A0 = (scanl (x,y.e1) v A)[1..n] in! \\
            && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
            && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A! \\
            && !let A3 = scanr * 1 A1! \\
            && !<reduce (x,y.e1) v A, fun (x1,...,xn,z) ->! \\
            && !let y1,B = map2 (x,y. x*y*z) (1::A2) A3 in! \\
            && !Y2(x1,...,xn,y1,B)>! \\
\end{tabular}
\end{center}

@@

\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
    &=& !let b=e1 in!   \\
    && !<if b then e2 else e3, fun (x1,...,xn,z) ->! \\
    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e2)(x1,...,xn,z)+!\\
    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e3)(x1,...,xn,z))>!
\end{tabular}

@@

\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
    &=& !let b=e1 in!   \\
    && !<if b then e2 else e3, fun (x1,...,xn,z) ->! \\
    && !Y((x1,...,xn)!$\widehat{+}$($\grad_\Gamma$!(e2)*b!$\widehat{+}\grad_\Gamma$!(e3)*(1-b))*z)>!\\
\end{tabular}

This time, there is a linear usage of the continuation variable !Y! (and of the variables !x1,...xn! too).

@@

A ground type $G$ is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context $\Gamma=x1:G1,...,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.

@@

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A1 \times ... \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times ... \times (An \otimes A)$ \\
    $\Array{A1} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times...\times An$ as expected.
TODO: define *, finish D(reduce)

@@

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !map2 (x,y.e1: G) e2 e3!) &=&  
    !let A,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y2 = !$\directD{\rho}{\Gamma,A}{Y1}$!(e3) in! \\
    && !let C=!$\grad_{\Gamma}$!e1 in!\\
    && !<map2 (x,y.e1) A B, fun (x1,...,xn,Z) -> !\\
    && !Y2( (x1,...,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e1)[a/x]!$\widehat{*}$!b) B Z )>!\\
    \end{tabular}
    \end{center}

\begin{center}
\begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
        && !let A0 = scanl (x,y.e1) y1 A in! \\
        && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
        && !let A2 = shift1 (map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A)! \\
        && !let A3 = scanr * 1 A1! \\
        && !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
        && !Y((x1,...,xn),0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

>>>>

We can now compute the reverse-mode transformation as follows.
\begin{center}
\begin{tabular}{r c l}
$\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
&=& $\ldots$ as before up until !A3! \\
&& !let B0 = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A0 A) in!\\
&& !let B1 = scanr !$1_\Gamma$ $\widehat{\times}$ !B0 in! \\
&& !let B2 = reduce !$\widehat{+}$ $0_\Gamma$! B1 in! \\
&& !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
&& !Y((x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!B2,0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

@@

Finally, the former point implies we need a few more array manipulations. 
We write !c::[v1,$\ldots$,vn]! to be the array ![c,v1,$\ldots$,vn]!. 
Given an array ![v1,$\ldots$,vn]! of size $n>0$, we write !let x,A=[v1,$\ldots$,vn]! to mean 
that !x=v1! and !A=[v2,$\ldots$,vn]!. The reverse derivative of fold left is then

TODO: reduce recomputed after scanl can be fixed here, explain let v,B notation somewhere
\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e1) e2 e3!) &=&
            !let v,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
            && !let A,Y2 = !$\directD{\rho}{\Gamma,v}{Y1}$!(e3) in! \\
            && !let A0 = (scanl (x,y.e1) v A)[1..n] in! \\
            && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
            && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A! \\
            && !let A3 = scanr * 1 A1! \\
            && !<reduce (x,y.e1) v A, fun (x1,$\ldots$,xn,z) ->! \\
            && !let y1,B = map2 (x,y. x*y*z) (1::A2) A3 in! \\
            && !Y2(x1,$\ldots$,xn,y1,B)>! \\
\end{tabular}
\end{center}

@@

\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
    &=& !let b=e1 in!   \\
    && !<if b then e2 else e3, fun (x1,$\ldots$,xn,z) ->! \\
    && !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e2)(x1,$\ldots$,xn,z)+!\\
    && \quad!(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e3)(x1,$\ldots$,xn,z))>!
\end{tabular}

@@

\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}(\Gamma \vdash$ !if e1 then e2 else e3!) 
    &=& !let b=e1 in!   \\
    && !<if b then e2 else e3, fun (x1,$\ldots$,xn,z) ->! \\
    && !Y((x1,$\ldots$,xn)!$\widehat{+}$($\grad_\Gamma$!(e2)*b!$\widehat{+}\grad_\Gamma$!(e3)*(1-b))*z)>!\\
\end{tabular}

This time, there is a linear usage of the continuation variable !Y! (and of the variables !x1,$\ldots$xn! too).

@@

A ground type $G$ is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context $\Gamma=x1:G1,\ldots,xn:Gn$ is interpreted as $\bigoplus_{1\leq i\leq n}Ai$, where $Ai\defeq\sem{Gi}$.
The denotation of the gradient of a term $\Gamma \vdash$!e: !$G$ at a point is then a matrix, more precisely an element of $()\bigoplus_{1\leq i\leq n}Ai)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}Ai\otimes A$.

@@

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A1 \times \ldots \times An)\otimes A$ & $\defeq$ & $(A1\otimes A) \times \ldots \times (An \otimes A)$ \\
    $\Array{A1}{n} \otimes A$ & $\defeq$ & $\Array{A1\otimes A}{n}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A1\times\ldotstimes An$ as expected.
TODO: define *, finish D(reduce)

@@

\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !map2 (x,y.e1: G) e2 e3!) &=&  
    !let A,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y2 = !$\directD{\rho}{\Gamma,A}{Y1}$!(e3) in! \\
    && !let C=!$\grad_{\Gamma}$!e1 in!\\
    && !<map2 (x,y.e1) A B, fun (x1,$\ldots$,xn,Z) -> !\\
    && !Y2( (x1,$\ldots$,xn)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e1)[a/x]!$\widehat{*}$!b) B Z )>!\\
    \end{tabular}
    \end{center}

\begin{center}
\begin{tabular}{r c l}
        $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
        &=&  !let y1,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
        && !let A,Y2 = !$\directD{\rho}{\Gamma,y1}{Y1}$!(e3) in! \\
        && !let A0 = scanl (x,y.e1) y1 A in! \\
        && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
        && !let A2 = shift1 (map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A)! \\
        && !let A3 = scanr * 1 A1! \\
        && !<reduce (x,y.e1) y1 A, fun (x1,$\ldots$,xn,z) ->! \\
        && !Y((x1,$\ldots$,xn),0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

### 2021-07-05 rad-icfp-2021/purerad.tex
minor syntax

The idea is that $\rho$ represents the return type of the derivative part, which should be !A$_{1} \times \ldots \times$ A$_n$! 
if we want the whole gradient of a term !e! in context !$\Gamma = \;$x$_{1}$:A$_{1}$,...,x$_n$:A$_n$!.
The subscript $\Gamma$ denotes the current context, 
which is locally augmented for instance when differentiating a !let! rule.
For non-unary operations, we differentiate the arguments from left to right and add their derivatives to the current stack, 
which is modeled by the continuation $Y$. 
Importantly for performance, each continuation variable $Y$ is only used once.

@@

By inspection in Figure~\ref{fig:direct_diff_macro}, 
we see that there is a linear usage of  each continuation $Y$ 
and that it is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,...,xn) -> Y(e1,...,en)! 
where the !ei! are !xi! except for at most $k$ (independant of $n$) terms.
In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x$_{1}$,...,x$_n$) -> (fun (y$_{1}$,...,yn) -> (f$_{1}$,...,f$_n$))(e$_{1}$,...,e$_n$)! 
$\transto$ 
!fun (x$_{1}$,...,x$_n$) -> let y$_{1}$,...,y$_n$ = e$_1$,...,e$_n$ in (f$_{1}$,...,f$_n$)!
$\transto$
!fun (x$_{1}$,...,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],...,f$_n$[e$_n$/y$_n$])!

>>>>

The idea is that $\rho$ represents the return type of the derivative part, which should be !A$_{1} \times \ldots \times$ A$_n$! 
if we want the whole gradient of a term !e! in context !$\Gamma = \;$x$_{1}$:A$_{1}$,$\ldots$,x$_n$:A$_n$!.
The subscript $\Gamma$ denotes the current context, 
which is locally augmented for instance when differentiating a !let! rule.
For non-unary operations, we differentiate the arguments from left to right and add their derivatives to the current stack, 
which is modeled by the continuation $Y$. 
Importantly for performance, each continuation variable $Y$ is only used once.

@@

By inspection in Figure~\ref{fig:direct_diff_macro}, 
we see that there is a linear usage of  each continuation $Y$ 
and that it is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
where the !ei! are !xi! except for at most $k$ (independant of $n$) terms.
In fact we have $k=2$, except for the !map2! case which requires a more detailed analysis (see Section~\ref{sec:complexity}).

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x$_{1}$,$\ldots$,x$_n$) -> (fun (y$_{1}$,$\ldots$,y$_n$) -> (f$_{1}$,$\ldots$,f$_n$))(e$_{1}$,$\ldots$,e$_n$)! 
$\transto$ 
!fun (x$_{1}$,$\ldots$,x$_n$) -> let y$_{1}$,$\ldots$,y$_n$ = e$_1$,$\ldots$,e$_n$ in (f$_{1}$,$\ldots$,f$_n$)!
$\transto$
!fun (x$_{1}$,$\ldots$,x$_n$) -> (f$_{1}$[e$_{1}$/y$_{1}$],$\ldots$,f$_n$[e$_n$/y$_n$])!

### 2021-07-05 rad-icfp-2021/unf.tex
minor syntax

The types are the same as in the source language. 
A context $\Gamma=\{$!x$_{1}$: T$_{1}$,...,x$_n$:T$_n$!$\}$ can always be seen as a type !T$_{1}$$\times$...$\times$ T$_n$!.
UNF is very syntactic and does not really use variables. 
In fact it consists only of constants and composition. 
For this reason, we denote contexts as types in the indices of the terms, and $\times$ is used for context extension.
But we do emphasize that we mean contexts and the actual variables are remembered and
this will be key when compiling to the target language after differentiation.

The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
By $\reals^{\times n}$ we mean the product $\reals\times...\times\reals$ of $n$ factors $\reals$.

>>>>

The types are the same as in the source language. 
A context $\Gamma=\{$!x$_{1}$: T$_{1}$,$\ldots$,x$_n$:T$_n$!$\}$ can always be seen as a type !T$_{1}$$\times\ldots\times$ T$_n$!.
UNF is very syntactic and does not really use variables. 
In fact it consists only of constants and composition. 
For this reason, we denote contexts as types in the indices of the terms, and $\times$ is used for context extension.
But we do emphasize that we mean contexts and the actual variables are remembered and
this will be key when compiling to the target language after differentiation.

The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
By $\reals^{\times n}$ we mean the product $\reals\times\ldotstimes\reals$ of $n$ factors $\reals$.
